Expectations bias moral evaluations

### 2018-07-03 paper/paper.tex
updating pre-submission

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{People living in geo-politically unstable regions or in the developing
world are often those who are most affected by terrorism, famine, and
natural disasters, and thus are the very people in greatest need of
assistance and concern from the world at-large. However, for these very
reasons, it is often unsurprising when harm befalls them. Two
experiments confirm this intuition: we found that people's expectations
bias their reactions to morally-significant events--people are more
upset by surprising harms than more expected harms. This bias threatens
to reduce observer's concern for victims most in need, making people
less likely to donate time or money to aid victims, or to take political
action to prevent further harms.}

\authorcontributions{Please provide details of author contributions here.}

\authordeclaration{Please declare any conflict of interest here.}

@@

When we learn of tragic events--a friend's loss of a loved-one, a
violent killing, or a terrorist attack--we are mournful, concerned,
angry, and driven to action. But our responses to these events differ
depending on the context (1). In particular, our expectations play an
important role in our reactions to events. This seems true of both the
mundane--when a friend's loved-one passes, we often ask, ``was it
sudden?''--the dramatic--a murder in a sleepy town prompts a very
different reaction than a similar crime in a major city--and even the
catastrophic: acts of terrorism seem to evoke more outrage, horror, and
concern when they occurs in peaceful rather than unstable regions. For
instance, the 2015 Paris attacks were felt around the world. Yet, most
of those mourning had little to say 15 hours earlier, when another
tragic attack killed at least 43 people in Beirut (2). Likewise, stories
of the Virginia Tech massacre, which killed 32 people, wildly
overshadowed coverage of attacks killing nearly 200 people in Iraq--one
of the bloodiest days since the 2003 invasion (3). In each of these
situations, the surprise elicited by an event appears to affect people's
reactions to the event. In some contexts, harm is tragic and in others,
``these things happen.''

Although this pattern seems so intuitive as to be obvious, no reasonable
code of ethics would justify this pattern of judgment, nor do any moral
psychological frameworks explicate the role of expectations in moral
evaluations. Outside of the domain of morality, however, everyday
experiences and controlled experiments show that people's evaluations of
events unambiguously depend on their expectations about those events.
There is a special thrill to having one's expectations exceeded and
disappointment when events fail to meet expectations: we root for the
underdog, hold surprise parties, and foreshadow bad news to ease its
delivery (4). Indeed, in a controlled study, Mellers and colleagues (5)
found that expectations influenced affective reactions during a gambling
task: Given a gamble with a 10\% chance to win \$30 and 90\% chance to
win \$0, participants felt little disappointment with the \$0 outcome,
but were considerably more excited when they won \$30. Conversely, given
a gamble with a 90\% chance of winning \$60 and 10\% chance of winning
\$0, participants were disappointed with the \$0 outcome and showed more
muted enjoyment of the \$60 outcome. In fact, in gambles similar to
these, people were happier with the smaller unexpected gain than with
the larger but more expected gain (6).

Applied to moral issues, the influences of expectations on evaluation
would be a dangerous bias. Tragedies are tragic, whether expected or
not, and failure to recognize that could lead to tacit endorsement of
the pain and suffering of others wherever pain and suffering are the
status quo. Do expectations directly bias reactions to morally
consequential events, as intuition and anecdote seem to suggest? To
demonstrate the plausibility of this suggestion, we develop a
theoretical account of how expectations influence evaluations, not as a
strange quirk or uniquely human bias, but instead as a fundamental
component of how people evaluate events.

Most events of consequence consist of changes--a transition from one
state to another. For this reason, we construe event evaluation as a
comparison between the states of the world before and after the event.
Comparison plays a crucial role in many theories of decision-making, and
especially in theories aimed at explaining the role of expectations in
evaluation (4, 5, 7, 8). For instance, in Prospect Theory (9, 10), the
utility of a prospect is evaluated by comparison to a current reference
point. Our proposal is similar: we argue that expectations can set the
reference points against which people compare future outcomes, so that
assigning value to an event consists in comparing the state of the world
following the event to the state of the world just prior to the event.
As knowledge of the world is uncertain, our understanding of these
states is probabilistic--that is, our model for the state of the world
is just our expectations about what has happened, what is happening, and
what will happen in the future. One direct implication of this proposal
is that the evaluation of an event is linked to what is learned about
the world as a result of that event's occurrence. That is, we evaluate
events positively or negatively to the extent that they inform our
understanding of the world in positive or negative ways.

This account thereby provides an information-theoretic route to
understanding how expectations directly influence evaluation. A
fundamental insight of Information Theory is that the information
carried by an event is a function of its prior probability; low
probability events carry more information than high probability events
(11). We propose that people learn more about the state of the world
when their expectations are violated by shocking world events as
compared to when they are affirmed by less surprising events. For
example, if a bombing occurs in Paris--a less expected location--rather
than Lebanon--a more expected location--we learn that the world is more
dangerous than we had previously believed. This suggestion is novel in
the moral domain, but these points are widely accepted elsewhere. For
instance, expectations have long been recognized as fundamental to
models of human and animal learning (12). Similarly, science progresses
most when new findings impugn widely-accepted theories or when a
theory's extreme predictions are validated.

@@

\begin{equation}
E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
\end{equation}

Given our construal of event evaluation as the comparison between states
of the world following and prior to the event, we define the value
assigned to the signalling event, \(V(S)\), as:

@@

The design of these studies was chosen with special consideration to two
theoretical points. First, the information-theoretic account we have
described makes a solely directional prediction--all else being equal,
more surprising events will elicit stronger reactions than less
surprising events. Asking participants to compare two similar events
allowed us to test this directional prediction. Second, it is the
informativeness of events, determined by one's prior expectations of the
event, that we predict will shape evaluations. Thus, reasoners should
not have to be induced to hold expectations experimentally (5), but
instead their own prior beliefs will be sufficient to shape their
expectations and affect their evaluations. For this reason, we
manipulated the prior probability of events by describing two otherwise
identical events occurring in different contexts. This allowed us to
test the role of expectations based on world knowledge, as well as
providing a more naturalistic set of items.

@@

Consistent with our predictions, manipulating expectations consistently
shaped evaluations across a range of morally-significant events. We
found that people viewed unexpected harmful events as more upsetting
than expected harmful events, even when the harm a victim suffered was
the same. For instance, people judged that a robbery at a clothing store
(unexpected) was more upsetting than a robbery at a convenience store
(expected). Furthermore, these differences were large, systematic, and
robust. Although there was variability in the situations that elicited
the largest differences, our statistical approach treated trial as a
random effect (see Methods and Results), allowing us to generalize
conclusions to items we did not test (14).

@@

Yet, however well-grounded the underlying principles, there's a clear
moral danger whenever expectations influence our moral concerns. For
instance, ``victim blaming'' describes the tendency to shift blame away
from aggressors and onto victims of crimes like sexual assault (15) and
police violence. Work by Janoff-Bulman and colleagues (16) has shown
that merely increasing the perceived probability of a victim suffering a
sexual assault induced participants to blame the victim for her assault.
People's expectations may explain the flip side of victim blaming,
whereby those who victim blame also tend to judge transgressors less
harshly: When people perceive a victims' actions as increasing the
probability of their eventual assault, they may perceive an aggressor's
actions as producing a smaller change in utility, thus warranting weaker
condemnation.

The influence of expectations on moral evaluations poses perhaps the
greatest threat when considering our reactions to humanitarian
challenges brought on by poverty or regional instability. In these
cases, the influence of expectations on moral evaluations threatens to
impose vicious and morally pernicious cycles. For instance, people
living in geo-politically unstable regions or in the developing world
are often those who are most affected by terrorism, famine, and natural
disasters, and are the very people in greatest need of assistance and
concern from the world at-large. However, for these very reasons, it is
often unsurprising when harm befalls people living in these
circumstances. Thus the influence of expectations on moral evaluations
threatens to reduce observer's concern for these victims most in need,
making people less likely to donate time or money to aid victims, or to
take political action to prevent further harms. The present research
identifies this potentially pernicious moral bias and grounds it in
information theoretic principles. Hopefully, by further understanding
these cognitive dynamics, psychologists and policymakers can begin to
find ways to reduce or counteract people's tendency to ignore the plight
of those in most need of attention.

\section*{Materials and Methods}\label{materials-and-methods}
\addcontentsline{toc}{section}{Materials and Methods}

We conducted two preregistered studies examining the role of
expectations in moral evaluations using two different judgment tasks
(\url{https://osf.io/5qnf7/}). Each study consisted of multiple trials
where participants were presented brief descriptions of two events.
Their task was to decide which of the two events was more upsetting. In
Study 1, they made a two-alternative forced-choice, and in Study 2 they
made their judgments using a five-point rating scale. Both studies
consisted of 16 experimental trials, five ``equivalent'' filler trials,
and five ``non-equivalent'' filler trials.

@@

Applying the expectation of conditional expectation, we can write:

\begin{equation}
V(s=1) = \sum_x u(x)p(x|s=1) - \sum_x \sum_s u(x)p(x|s)p(s)
\end{equation}

We then perform some algebra. We break out the summation over \(S\) (6),
pull the multipliers out of the last two sums (7), rewrite \(p(s=0)\) in
terms of \(p(s=1)\) (8), and factor to combine the first two sums (9).

\begin{align}
V(s=1) &= \sum_x u(x)p(x|s=1) - \sum_x u(x)p(x|s=1)p(s=1) - \sum_x u(x)p(x|s=0)p(s=0) \\
 &= \sum_x u(x)p(x|s=1) - p(s=1)\sum_x u(x)p(x|s=1) - p(s=0) \sum_x u(x)p(x|s=0) \\
 &= \sum_x u(x)p(x|s=1) - p(s=1)\sum_x u(x)p(x|s=1) - (1-p(s=1)) \sum_x u(x)p(x|s=0) \\
 &=(1-p(s=1))\sum_x u(x)p(x|s=1) - (1-p(s=1)) \sum_x u(x)p(x|s=0)
\end{align}

Finally, combining both sums by their common factor yields (10):

@@

Thus, it can be seen under this formalization that the role of
expectations in the evaluation of events is fundamental--it does not,
for instance, depend on any specifics concerning the utility function
over states of the world. Instead, this conclusion depends on just two
assumptions: first, that events are evaluated by comparing the utilities
of the states of the world prior to and following the event; and second,
that knowledge of those states of the world is uncertain.

We have appealed to the informativeness of an event to illustrate
intuitively how expectations should influence evaluations. More expected
events carry less information, and thereby engender weaker reactions.
Though our model of event evaluation is not an information-theoretic
model in a direct sense, this interpretation still applies: the
information carried by observing \(S\) is a function of that
observation's prior probability, which we have in turn shown determines
its evaluation. Also relevant, although not addressed here, is the
mutual information between \(X\) and \(S\)--the extent to which \(S\)
informs \(X\).

>>>>

\documentclass[9pt,twocolumn,twoside,]{pnas-new}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{In two behavioral studies, we found that people's expectations biased
their reactions to morally-significant events--people are more upset by
surprising harms than more expected harms. This bias appears deeply
pernicious: because it is often unsurprising when people living in
poverty, dangerous neighborhoods, or unstable regions become victims of
crime, natural disasters, or terrorism, observers are likely to
experience reduced concern for these victims. Yet people living in
disadvantaged circumstances are those most in need of assistance and
concern from the world at-large. By blunting people's concern for those
victims who are most in need, the bias we have uncovered threatens to
increase their suffering and further sustain inequality.}

\authorcontributions{Both authors jointly conceived of the project and theory, designed and
conducted experiments, and analyzed results. D.P. developed the formal
aspects of the theory.}

\authordeclaration{The authors declare no conflicts of interest.}

@@

When we learn of tragic events--a friend's loss of a loved-one, a
violent killing, or a terrorist attack--we are mournful, concerned,
angry, and driven to action. But our responses to these events differ
depending on the context in which these events occur (1). In particular,
our expectations play an important role in our reactions to events. This
seems true of both the mundane--when a friend's loved-one passes, we
often ask, ``was it sudden?''--the dramatic--a murder in a sleepy town
prompts a very different reaction than a similar crime in a major
city--and even the catastrophic: acts of terrorism seem to evoke more
outrage, horror, and concern when they occurs in peaceful rather than
unstable regions. For instance, the 2015 Paris attacks were felt around
the world. Yet, most of those mourning had little to say 15 hours
earlier, when another tragic attack killed at least 43 people in Beirut
(2). Likewise, stories of the Virginia Tech massacre, which killed 32
people, wildly overshadowed coverage of attacks killing nearly 200
people in Iraq--one of the bloodiest days since the 2003 invasion (3).
In each of these situations, the surprise elicited by an event appears
to affect people's reactions to the event. In some contexts, harm is
tragic and in others, ``these things happen.''

Although this pattern seems so intuitive as to be obvious, no reasonable
code of ethics would justify this pattern of judgment, nor do any moral
psychological frameworks explicate the role of expectations in moral
evaluations. Outside of the domain of morality, everyday experiences and
controlled experiments show that people's evaluations of events
unambiguously depend on their expectations about those events. There is
a special thrill to having one's expectations exceeded and
disappointment when events fail to meet expectations: we root for the
underdog, hold surprise parties, and foreshadow bad news to ease its
delivery (4). Indeed, in a controlled study, Mellers and colleagues (5)
found that expectations influenced affective reactions during a gambling
task: Given a gamble with a 10\% chance to win \$30 and 90\% chance to
win \$0, participants felt little disappointment with the \$0 outcome,
but were considerably more excited when they won \$30. Conversely, given
a gamble with a 90\% chance of winning \$60 and 10\% chance of winning
\$0, participants were disappointed with the \$0 outcome and showed more
muted enjoyment of the \$60 outcome. In fact, in gambles similar to
these, people were happier with the smaller unexpected gain than with
the larger but more expected gain (6).

The influence of expectations on moral evaluations threatens to produce
negative social consequences. Tragedies are tragic, whether expected or
not, and failure to recognize that could lead to tacit endorsement of
the pain and suffering of others wherever pain and suffering are the
status quo. Do expectations directly bias reactions to morally
consequential events, as intuition and anecdote seem to suggest? To
demonstrate the plausibility of this suggestion, we develop a
theoretical account of how expectations influence evaluations, not as a
strange quirk or uniquely human bias, but instead as a fundamental
component of how people evaluate events.

Most events of consequence consist of changes--a transition from one
state to another. For this reason, we construe event evaluation as a
comparison between the states of the world before and after the event.
Comparison plays a crucial role in many theories of decision-making, and
especially in theories aimed at explaining the role of expectations in
evaluation (4, 5, 7, 8). For instance, in Prospect Theory (9, 10), the
utility of a prospect is evaluated by comparison to a current reference
point. Our proposal is similar: we argue that expectations can set the
reference points against which people compare future outcomes, so that
assigning value to an event consists in comparing the state of the world
following the event to the state of the world just prior to the event.
We further assume that people's knowledge of the world is uncertain, so
we represent the states of the world probabilistically. This means that
a person's model for the state of the world is just their expectations
about what has happened, what is happening, and what will happen in the
future. One direct implication of this proposal is that the evaluation
of an event is linked to what is learned about the world as a result of
that event's occurrence. That is, we evaluate events positively or
negatively to the extent that they inform our understanding of the world
in positive or negative ways.

One striking implication of this account is that it provides an
information-theoretic route to understanding how expectations directly
influence evaluation. A fundamental insight of Information Theory is
that the information carried by an event is a function of its prior
probability; low probability events carry more information than high
probability events (11). We propose that people learn more about the
state of the world when their expectations are violated by shocking
world events as compared to when they are affirmed by less surprising
events. For example, if a bombing occurs in Paris--a less expected
location--rather than Lebanon--a more expected location--we learn that
the world is more dangerous than we had previously believed. This
suggestion is novel in the moral domain, but these points are widely
accepted elsewhere. For instance, expectations have long been recognized
as fundamental to models of human and animal learning (12).

@@

\begin{equation*}
E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
\end{equation*}

Our construal of event evaluation is as a comparison between states of
the world following and prior to an event. Consequently, we define the
value assigned to the signalling event, \(V(S)\), as:

\begin{equation*}
V(S) = E_{X|S}[u(X)] - E_X[u(X)]
\end{equation*}

@@

The design of these studies was chosen with special consideration to two
theoretical points. First, the information-theoretic account we have
described makes a solely directional prediction--all else being equal,
more surprising events will elicit stronger reactions than less
surprising events. Asking participants to compare two similar events
allowed us to test this directional prediction. Second, it is the
informativeness of events, determined by one's prior expectations of the
event, that we predict will shape evaluations. Thus, reasoners should
not have to be induced to hold expectations experimentally (5), but
instead their own prior beliefs will be sufficient to shape their
expectations and affect their evaluations. For this reason, we
manipulated the prior probability of events by describing two otherwise
identical events occurring in different contexts. We explicitly chose
contexts that differed in ways that could not be accounted for by
construal-level explanations ((1)) or explanations based on biases
towards one's ingroup (\textbf{???}). This allowed us to test the role
of expectations based on world knowledge, as well as providing a more
naturalistic set of items.

@@

Consistent with our predictions, manipulating expectations consistently
shaped evaluations across a range of morally-significant events. We
found that people viewed unexpected harmful events as more upsetting
than expected harmful events, even when the harm a victim suffered was
the same. For instance, people judged that a robbery at a clothing store
(unexpected) was more upsetting than a robbery at a convenience store
(expected). Furthermore, these differences were large, systematic, and
robust. Although there was variability in the situations that elicited
the largest differences, our statistical approach treated trial as a
random effect (see Methods and Results), allowing us to generalize our
conclusions to items we did not test (14).

@@

Yet, however well-grounded the underlying principles, there's a clear
moral danger whenever expectations influence our moral concerns (15,
16). These effects pose a significant threat when considering our
reactions to humanitarian challenges brought on by poverty or regional
instability. In these cases, the influence of expectations on moral
evaluations threatens to impose vicious and morally pernicious cycles.
For instance, people living in geo-politically unstable regions or in
the developing world are often those who are most affected by terrorism,
famine, and natural disasters, and are the very people in greatest need
of assistance and concern from the world at-large. However, for these
very reasons, it is often unsurprising when harm befalls people living
in these circumstances. Thus the influence of expectations on moral
evaluations threatens to reduce observer's concern for these victims
most in need, making people less likely to donate time or money to aid
victims, or to take political action to prevent further harms. The
present research identifies this potentially pernicious moral bias and
grounds it in information theoretic principles. Hopefully, by further
understanding these cognitive dynamics, psychologists and policymakers
can begin to find ways to reduce or counteract people's tendency to
ignore the plight of those in most need of attention.

\matmethods{

We conducted two preregistered studies examining the role of
expectations in moral evaluations (\url{https://osf.io/5qnf7/}). Each
study consisted of multiple trials where participants were presented
brief descriptions of two events. Their task was to decide which of the
two events was more upsetting. In Study 1, they made a two-alternative
forced-choice, and in Study 2 they made their judgments using a
five-point rating scale. Both studies consisted of 16 experimental
trials, five ``equivalent'' filler trials, and five ``non-equivalent''
filler trials.

@@

Our goal is to express this relation in terms of \(p(S)\). First, we
manipulate \(E_X[u(X)]\) by applying the expectation of conditional
expectation and some algebra.

\begin{align*}
\sum_x u(x)p(x) & = \sum_s u(x)p(x|s)p(s) \\
 &= \begin{aligned}[t]
    & \sum_x u(x)p(x|s=1)p(s=1) - \\
    & \quad \sum_x u(x)p(x|s=0)p(s=0) \end{aligned}\\
 &= \begin{aligned}[t]
    & p(s=1)\sum_x u(x)p(x|s=1) -\\
    & \quad p(s=0) \sum_x u(x)p(x|s=0) \end{aligned}\\
 &= \begin{aligned}[t]
    & p(s=1)\sum_x u(x)p(x|s=1) - \\
    & \quad (1-p(s=1)) \sum_x u(x)p(x|s=0) \end{aligned}\\
\end{align*}

Substituting this quantity back into Equation 2, we obtain:

\begin{align*}
V(s=1) = & \sum_x u(x)p(x|s=1) - \\ & p(s=1)\sum_x u(x)p(x|s=1) -\\ & (1-p(s=1)) \sum_x u(x)p(x|s=0)
\end{align*}

Finally, we factor the first two sums and combine them with the last sum
to yield:

@@

Thus, it can be seen under this formalization that the role of
expectations in the evaluation of events is fundamental in the sense
that it does not depend on any specifics concerning the utility function
over states of the world. Instead, this conclusion depends on just two
assumptions: first, that events are evaluated by comparing the utilities
of the states of the world prior to and following the event; and second,
that knowledge of those states of the world is uncertain.

We have appealed to the informativeness of an event to illustrate
intuitively how expectations should influence evaluations. More expected
events carry less information, and thereby engender weaker reactions.
Though our model of event evaluation is not an information-theoretic
model in a direct sense, this interpretation still applies: the
information carried by observing \(S\) is a function of that
observation's prior probability, which we have in turn shown determines
its evaluation.
}
